Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-09-09 latex/flat.tex
flat.tex

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

@@

\section{Introduction}\label{section:intro}
The practice of experimental design entails specifying every detail of
running an experiment \cite{chaloner1995}. At its simplest form,
experimental design involves choosing a set of measurements to be
taken. For example, consider the process of searching for oil:
measurements require digging holes in the ground and a \emph{design}
simply indicates which holes should be dug
\cite{horesh2008borehole}. Since measurements are expensive, only a
limited number of them can be taken, and thus carefully choosing
measurement locations is crucial.

@@

Inverse problems are ubiquitous in science and engineering. For
example, in electrical impedance tomography (EIT) we seek to infer the
structure of the inside of a human body. Inference is conducted based
on observations of electrical current impedance measured at specific
electrode locations on the skin, leveraging known equations of
electric current flow \cite{horesh2010impedance}. Magnetic Resonance
Imaging (MRI) also entails solving an inverse problem. There,
radio-frequency pulses are sent through the human body in the presence
of a strong magnetic field. The response of one's body contents to
these radio-frequency pulses is measured, allowing a radiologist to
view the internals of a body in a noninvasive manner
\cite{horesh2008mri}. An inverse problem also arises in oil and gas
exploration, where acoustic waves are sent through "boreholes" ---
deep cylindrical wells drilled into the ground. Data of travel and
return times of the acoustic waves are recorded. Wave travel and
return time is influenced by the properties of the subsurface
materials, such as density, elasticity, and the presence of fluids or
voids. Combining travel time data with a geophysical models of
contents of the earth's crust facilitates reconstructing the structure
of the subsurface in a process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola2005}. In many inverse problems the
goal is to infer some numerical parameter. However, in the formulation
we consider in this study, as well as in the examples above, the goal
is to conduct inference over some \emph{function} over $\Omega$, where
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain of
interest.

\subsection{D-optimal Designs}\label{subsec:D}
In all of the examples of inverse problems we have seen, as well as in
many other applications, only a limited number of measurements are
allowed. For example, in borehole tomography, a measurement involves
drilling a deep well in the ground --- a very costly endeavor. Even in
a seemingly simpler application such as MRI, a patient cannot occupy
the MRI machine for too long, hence only a limited number of
measurements can be taken. Therefore, measurements should be chosen to
most benefit the user. Typically, a user will consider some utility,
called a \emph{design criterion} and take measurements that maximize
this utility. Two of the most widely recognized and extensively
studied design criteria are the A- and \emph{D-optimality} criteria.

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Leibler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}. Utilizing KL divergence for quantifying differences
between distributions is standard practice and we will not motivate it
further here. For a discrete parameter $\param$ and data $\data$, the
KL divergence is defined as
\begin{equation}\label{eq:basic_KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted, so we average over it to define the D-optimality criterion
as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ].
\end{equation*}
We refer to a set of measurements that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

For a linear model with Gaussian prior in finite dimensions, a
D-optimal design minimizes the determinant of the posterior covariance
\cite{chaloner1995}. In Section \ref{subsec:D_optimal_design} we give
a more general definition of the D-optimality criterion that applies
to arbitrary measures. We also show how the D-optimality criterion
generalizes to linear models over infinite-dimensional Hilbert spaces
(i.e.~function spaces).

\subsection{A toy model: the 1D heat equation}\label{subsec:toy}
For concreteness, let us now consider a toy inverse problem: inferring
the initial condition for partial differential equation known as the
\emph{heat equation} in one dimension (1D heat equation
henceforth). Readers less familiar with partial differential equations
can think of the following setup: before us there is a perfectly
insulated metal rod. At each point on the rod the temperature is
different and unknown, and we call this temperature distribution the
\emph{initial condition}. We wait for a short time $T$ and let the
heat dissipate a bit inside the rod. The heat equation determines the
heat distribution inside the rod at time $t=T$.

@@

Solving the 1D heat equation is straightforward. Any initial heat
distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition eq.~\eqref{eq:heat2} is a linear
combination of sines $\ev_n(x) = \sin(\pi n x)$, so $u_0 = \sum_{n\geq
  1} a_n \ev_n$ for some $\{a_n\}_{n\geq 1}$. Now, note that $\ev_n$
are in fact eigenvectors of the Laplacian: $\Delta \ev_n = -\pi^2
n^2\ev_n$. It should not be hard to believe that $u(\cdot, T) =
\sum_{n\geq 1} a_n \exp(-\pi^2 n^2T ) \ev_n$
\cite{renardy2006PDE}. Thus, eigenvectors of the linear operator that
describes the time evolution of heat are $\ev_n(x) = \sin(\pi n x)$
with corresponding eigenvalues
\begin{equation}\label{eq:decay}
  \exp(-\pi^2 n^2T ).
\end{equation}

In the \emph{inverse problem of the 1D heat equation}, our goal is to
infer the initial condition $u_0 = u(\cdot, 0)$ from observations of
the final state $u(\cdot, T)$. However, we are not able to measure
$u(\cdot, T)$ at every point in the domain $\Omega$. Rather, we take
some number of temperature measurements on the rod once our waiting
time $T$ has passed and try to infer $u(\cdot, 0)$ from these
measurements.

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, as implied by the squared
exponential decay of eigenvalues in eq.~\eqref{eq:decay} with
$n\to\infty$. See Supplementary movies S1 and S2 for an illustration
of this phenomenon (code generating these movies is located in module
\texttt{movies.py} in the accompanying
\href{https://github.com/yairdaon/OED}{repository}). For ill-posed
problems regularization is required, and we implement such
regularization via a fully Bayesian formulation of the inverse
problem.

We choose a prior for the initial condition $u_0 \sim \mathcal{N}(0,
(-\Delta)^{-1})$, with homogeneous Dirichlet boundary condition. This
choice of prior covariance operator ensures the posterior is
well-defined and prior realizations are well-behaved, see
\cite[Theorem 3.1 and Lemma 6.25]{Stuart10} for details. The second
ingredient of a fully Bayesian formulation of the inverse problem is
likelihood. We simply assume centered iid Gaussian measurement error
model, giving rise to a the widely applicable Gaussian likelihood
function.

\subsection{Measurement Clusterization}
Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorovDesignSpatialExperiments1996, nyberg2012, fedorov1997,
  Ucinski05, neitzel2019sparse}. This phenomenon is illustrated for
our toy inverse problem in Fig.~\ref{fig:clusterization_illustration},
where D-optimal measurement locations are shown for different numbers
of measurements. Notably, for six measurements, a D-optimal design
yields two sets of measurements that are identical. Following
\cite{Ucinski05}, we refer to this intriguing phenomenon as
\emph{measurement clusterization}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

@@

Clusterization should not be confused with either repetition nor with
replication, which are commonly viewed as beneficial and even
necessary aspects of an optimal design \cite{fisher1949design,
  morris2011, schafer2001replication}. For example,
\cite{fisher1949design} in his famous milk and tea experiment,
suggested that repetition is a "way of enlarging the experiment and,
thereby, increasing its sensitiveness". In another example,
\cite{fay2000rainfall} measured the effect of rainfall on grass growth
in plots of land. The experiment involved fifteen "rainfall
manipulation shelters", where "Four rainfall manipulation treatments
(three replicates) then were assigned to 12 of the plots". While it
seems reasonable for the researchers to replicate the phenomenon they
are trying to study, clusterization is different: a clustered design
in the rainfall experiment would imply the researchers should take a
repeated measurement \emph{on the same plot}, at the expense of
measuring grass growth in other plots. In sharp contrast to repetition
and replication, clusterization is highly nonintuitive and we will see
that its origins run considerably deeper.

Researchers of inverse problems widely agree that measurement
clusterization is undesirable
\cite{fedorovDesignSpatialExperiments1996, nyberg2012, fedorov1997,
  Ucinski05, neitzel2019sparse}, prompting the exploration of various
remedies to address this issue. One approach involves merging close
measurements \cite{fedorov1997}; however, this strategy merely
overlooks the phenomenon of measurement clusterization. An alternative
solution lies in \emph{clusterization-free design}s, where measurement
locations are deliberately chosen to be distant from one another. This
can be achieved by imposing distance constraints between measurements
or by introducing correlated errors that account for both observation
error and model misspecification \cite{Ucinski05}. For instance, in
the context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{nyberg2012}.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior covariance in
  observation space decay. This decay depends mostly on how ill-posed
  the problem at hand is and does not depend much on the prior. See
  Section \ref{section:vanishing} --- particularly, Theorem
  \ref{thm:char} and the discussion following it. We also show results
  of numerical experiments, where simulations of our model give rise
  to D-optimal designs that exhibit clusterization with high
  probability in Section \ref{subsec:lemma_sims}. Thus, given the
  genericity of our model, we expect measurement clusterization to be
  a generic and ubiquitous phenomenon.

@@

  The connection to Carath\'eodory's Theorem also utilizes the
  concentration of measurement effort on the abovementioned select set
  of prior eigenvectors. This concentration allows us to move from an
  infinite-dimensional setting to finite dimensions, where the
  conditions of Carath\'eodory's Theorem hold. We conclude that a
  D-optimal design arises by weighting a small number of
  measurements. If the number of allowed measurements is larger than
  the small number dictated by Carath\'eodory's Theorem, we have
  excessive weight on some measurements, which can be interpreted as
  clusterization.

@@

\subsubsection{Implications}
Our answer to Question \ref{q:generic} implies that encountering
clusterization should be expected in many different problems across
many different scientific fields. Researchers that encounter
clusterization should not be surprised or wary. In Our answer to
Question \ref{q:why}, we explain what our view of the cause of
clusterization is. It appears, the cause is generic: a D-optimal
design reduces uncertainty for a select set of prior covariance
eigenvectors --- those with the most prior uncertainty, i.e.~those the
practitioner cares about the most! We believe practitioners should not
try to avoid measurement clusterization. Rather, practitioners should
take repeated measurements (e.g.~in MRI and borehole tomography),
increase apparatus sensitivity (e.g.~in EIT), or take consecutive
measurements (e.g.~in the 1D heat equation). Overall, we show that
measurement clusterization is a natural and (almost) inevitable part
of Bayesian D-optimal designs (but see disclaimer below).

One interesting implication of the analysis presented here is that
clusterization can serve as an evidence to the number of relevant
eigenvectors. Since leading eigenvectors typically correspond to slow
variations in space and/or time, clusterization could be used to
estimate the number of relevant degrees of freedom, and even to reduce
the complexity of a computational model, e.g.~by dropping
discretization points.

It is important to note that we do not view clustered designs as
undesirable, nor do we believe it should be avoided at
all. Clusterization is a peculiar phenomenon and it is perfectly
reasonable for someone to argue against the D-optimality criterion
based on the fact that it results in clustered designs. We have seen,
however that there is a perfectly reasonable explanation for
clusterization. We have shown that clusterization is an inevitable
consequence of having a problem with some modes where uncertainty
decays faster than others.

Lastly, we believe that when clusterization arises, it should serve as
a warning sign to practitioners. In the inverse problem of the 1D heat
equation, clusterization occurs primarily because Laplacian
eigenvectors \emph{do not} decay in $\Omega$. Consequently, measuring
$u(x_1, T)$ at some point $x_1 \in \Omega$ provides information about
$u(x_2,T)$ for distant points $x_2 \in \Omega$. Intuitively, this
should not occur: for small $T$, the heat distribution at $x_1$ should
give very little knowledge on the heat distribution at $x_2$. This
behavior stems from a well-known property of the heat equation: it
allows information to spread \emph{instantly} across the computational
domain \cite{renardy2006PDE}. In reality, heat (and information)
propagate at finite speeds. Of course, the known physical barrier for
information spread is the speed of light, but we expect heat to spread
considerably slower: heating an Olympic pool at one end should have no
immediate effect on the temperature at the other end.

Our choice of prior is also a potential major contributor to
clusterization. Our choice of Gaussian prior similarly implies
information is shared between distant locations in $\Omega$. Thus, we
suggest refraining from choosing Gaussian priors with inverse
Laplacian covariance operators. Rather, non-Gaussian priors could be
employed instead \cite{hosseini2017, hosseini2019}.

We believe that the emergence of clusterization in this context is
thus non-physical, arising from the way the inverse problems we
consider are phrased. Clusterization therefore indicates that the
underlying mathematical / Bayesian model is overly permissive and
fails to capture crucial physical constraints of the problem. We
suggest that when clusterization occurs, practitioners should consider
alternative models where information is localized in space and travels
at finite speed in the medium. Such models may not only provide more
physically accurate and meaningful results but may also mitigate the
issue of clusterization.

\subsubsection{Other Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the space of \emph{observations}, see Section
\ref{section:vanishing} for a precise statement. This is in accordance
with previous work by \cite{koval2020}, who showed that A-optimal
designs are best constructed in the space of observations.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. Specifically, in its
current form, our model does not allow point evaluations. Thus, while
our model is generic enough to be analytically tractable, some may
argue that our model is too far removed from any real application. To
these claims I would answer that scientists have a long history of
studying models that are bare-bones simplifications of real systems,
e.g.~the Ising model \cite{cipra1987}, the Lorenz system \cite{brin},
the Lotka-Volterra equations \cite{logan2006}, the Carnot engine
\cite{kardar2007} and many others.
 \section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over Hilbert
spaces. Since we are ultimately interested in inferring a function
over some domain, we will keep in mind that these Hilbert spaces
should really be thought of as function spaces. A deeper treatment of
the foundations for inverse problems over function spaces can be found
in \cite{Stuart10}.

\subsection{Forward Problems}\label{subsec:abstract_OED}
In this section we give definitions and notations for forward problems,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

@@

Proposition \ref{prop:bigger_better} implies that it is a good idea to
bound the norm of measurements. If, for example, we can take
measurements in $\textup{span}\{\meas\}$ for some $\meas \neq 0$, then
the D-optimality criterion is unbounded, so a D-optimal design does
not exist. In contrast, in any real-life problem where sensors are
concerned, the norm of measurements recorded by sensors is always
one\footnote{Again, our analysis does not directly apply to point
evaluations. We just utilize point evaluations for motivation. We can
approximate point evaluations by e.g.~elements in $\hilo^*$ as long as
$\hilo$ is a function space, e.g.~$L^2(\Omega)$. In this case, for a
fixed apporximation of $\delta$ the norm of the corresponding
functional is constant (but $\neq 1$.}:

>>>>

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\raggedbottom

@@

\section{Introduction}\label{section:intro}
Experimental design is a systematic approach to specifying every
possible aspect of an experiment \cite{chaloner1995}. In this study,
we focus on experiments that involve measuring quantities arising from
naturally occurring processes. For us, therefore, experimental design
entails determining which specific measurements to record from a given
process. In a typical experiment the number of measurements that can
be recorded is limited by time, cost and other constraints. Therefore
selecting the right set of measurements to record is crucial as it
directly impacts the utility of the data collected, and consequently,
the validity of the conclusions drawn.

For example, consider the process of searching for oil: measurements
require digging deep holes in the ground, known as "boreholes". A
\emph{design} simply indicates which boreholes should be dug
\cite{horesh2008borehole}. Since digging boreholes is expensive, only
a limited number of them can be dug, and thus carefully choosing their
locations is crucial.

@@

Inverse problems are ubiquitous in science and engineering. For
example, in electrical impedance tomography (EIT) we seek to infer the
structure of the inside of a human body. Inference is conducted based
on observations of electrical current impedance measured at specific
electrode locations on the skin, leveraging known equations of
electric current flow \cite{horesh2010impedance}. Magnetic Resonance
Imaging (MRI) also entails solving an inverse problem. There,
radio-frequency pulses are sent through the human body in the presence
of a strong magnetic field. The response of one's body contents to
these radio-frequency pulses is measured, allowing a radiologist to
view the internals of a body in a noninvasive manner
\cite{horesh2008mri}. An inverse problem also arises in oil and gas
exploration, where acoustic waves are sent through "boreholes" ---
deep cylindrical wells drilled into the ground. Data of travel and
return times of the acoustic waves are recorded. Wave travel and
return time is influenced by the properties of the subsurface
materials, such as density, elasticity, and the presence of fluids or
voids. Combining travel time data with a geophysical model of the
contents of the earth's crust facilitates reconstructing the structure
of the subsurface in a process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola2005}. In many inverse problems the
goal is to infer some numerical parameter. However, in the formulation
we consider in this study, as well as in the examples above, the goal
is to conduct inference over some \emph{function} over $\Omega$, where
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain of
interest.

\subsection{D-optimal Designs}\label{subsec:D}
In all of the examples of inverse problems we have seen, as well as in
many other applications, only a limited number of measurements are
allowed. For example, in borehole tomography, a measurement involves
drilling a deep well in the ground --- a very costly endeavor. In
medical imaging applications like MRI, the time allotted for each
patient in the MRI machine limits the number of measurements that can
be taken during a single session. Therefore, measurements should be
chosen to extract as much information from the experiment. Typically,
a user will consider some utility, called a \emph{design criterion}
and take measurements that maximize this utility. Two of the most
widely recognized and extensively studied design criteria are the A-
and \emph{D-optimality} criteria.

In this study, we focus on the Bayesian D-optimality
criterion. Bayesian D-optimality carries a simple and intuitive
meaning: measurements chosen according to the D-optimality criterion
maximize the expected Kullback-Leibler (KL) divergence from posterior
to prior \cite{chaloner1995, AlexanderianGloorGhattas14,
  CoverThomas91}.
Recall that for a discrete parameter $\param$ and data $\data$, the KL
divergence is defined as
\begin{equation}\label{eq:basic_KL}
  D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) =  \sum_{\param} \log
  \frac{\Pr(\param|\data)}{\Pr(\param)} \Pr(\param|\data). 
\end{equation}
Of course, data $\data$ is not known before the experiment is
conducted. Hence, we average over $\data$ to define the D-optimality
criterion as:
\begin{equation*}
  \mathbb{E}_{\data}\left [ D_{KL}\left (\Pr(\param|\data)||\Pr(\param)\right ) \right ].
\end{equation*}
We refer to a set of measurements that maximizes the D-optimality
criterion as a \emph{D-optimal design}.

For a linear model with Gaussian prior in finite dimensions, a
D-optimal design minimizes the determinant of the posterior covariance
\cite{chaloner1995}. In Section \ref{subsec:D_optimal_design} we give
a more general definition of the D-optimality criterion that applies
to arbitrary measures. We also show how the D-optimality criterion
generalizes to linear models over infinite-dimensional Hilbert spaces
(e.g.~function spaces).

\subsection{A toy model: the 1D heat equation}\label{subsec:toy}
For concreteness, let us now consider a toy inverse problem: inferring
the initial condition for a partial differential equation known as the
\emph{heat equation} in one dimension (1D heat equation
henceforth). Readers less familiar with partial differential equations
can think of the following setup: before us there is a metal rod,
perfectly insulated except for its tips. At each point on the rod the
temperature is different and unknown, and we call this temperature
distribution the \emph{initial condition}. We wait for a short time
$T$ and let the heat dissipate a bit inside the rod. The heat equation
determines the heat distribution inside the rod at time $t=T$.

@@

Solving the 1D heat equation is straightforward. Consider an initial
heat distribution $u_0\in L^2([0,1])$ that satisfies the homogeneous
Dirichlet boundary condition eq.~\eqref{eq:heat2}. This initial
condition $u_0$ is a linear combination of sines $\ev_n(x) = \sin(\pi
n x)$, so $u_0 = \sum_{n\geq 1} a_n \ev_n$ for some $\{a_n\}_{n\geq
  1}$. Now, note that $\ev_n$ are in fact eigenvectors of the
Laplacian: $\Delta \ev_n = -\pi^2 n^2\ev_n$. It should not be hard to
believe that $u(\cdot, T) = \sum_{n\geq 1} a_n \exp(-\pi^2 n^2T )
\ev_n$ \cite{renardy2006PDE}. Thus, eigenvectors of the linear
operator that describes the time evolution of heat are $\ev_n(x) =
\sin(\pi n x)$ with corresponding eigenvalues
\begin{equation}\label{eq:decay}
  \exp(-\pi^2 n^2T ).
\end{equation}

In the \emph{inverse problem of the 1D heat equation}, our goal is to
infer the initial condition $u_0 = u(\cdot, 0)$ from noisy
observations of the final state $u(\cdot, T)$. However, we are not
able to measure $u(\cdot, T)$ at every point in the domain
$\Omega$. Rather, we take some number of temperature measurements on
the rod once our waiting time $T$ has passed and try to infer
$u(\cdot, 0)$ from these measurements.

The inverse problem of the 1D heat equation is difficult ("ill-posed")
since heat spreads in a diffusive manner and any roughness in the
initial condition is quickly smoothed, as implied by the squared
exponential decay of eigenvalues in eq.~\eqref{eq:decay} with
$n\to\infty$. See Supplementary movies S1 and S2 for an illustration
of this phenomenon\footnote{Code generating these movies is located in
module \texttt{movies.py} in the accompanying
\href{https://github.com/yairdaon/OED}{repository}.}. For ill-posed
problems regularization is required, and we implement such
regularization via a fully Bayesian formulation of the inverse
problem.

In order to find a fully Bayesian formulation for our inverse problem
we first need to specify a prior on \emph{functions} defined over
$\Omega = [0,1]$. For the sake of simplicity we would like to utilize
a Gaussian prior. While specifying the prior mean is easy --- we take
the zero function on $\Omega$ --- specifying a prior covariance is
more involved. In analogy with the finite dimensional case, we seek a
covariance \emph{operator} that is positive definite and imposes
sufficient regularity on functions over $\Omega$. Thus, we set our
prior for the initial condition $u_0 \sim \mathcal{N}(0,
(-\Delta)^{-1})$, where $\Delta$ is again defined with a homogeneous
Dirichlet boundary condition. The intuition here is that since
$\Delta$ is a differential operator, it is "roughing" (the opposite of
smoothing) and so $(-\Delta)^{-1}$ is smoothing. Prior realizations
are generated, in analogy to the finite dimensional case, by smoothing
white noise: $(-\Delta)^{-1/2}\mathcal{W}$, where $\mathcal{W}$ is
white noise \cite{rue2011}. Our choice of prior covariance operator
ensures the posterior is well-defined and prior realizations are
well-behaved, see Theorem 3.1 and Lemma 6.25 in \cite{Stuart10} for
details. The second ingredient of a fully Bayesian formulation of the
inverse problem is likelihood. We simply assume centered iid Gaussian
measurement error model, giving rise to a the widely applicable
Gaussian likelihood function.

Now that we have a fully Bayesian formulation of our toy inverse
problem, we would like to find a D-optimal design. Defining D-optimal
designs over function spaces is somewhat mathematically
involved. Thus, to keep our discussion here focused, we will explore
these technical details later in Section
\ref{subsec:D_optimal_design}. For now, it suffices to know that
D-optimal designs can be defined for our problem.

\subsection{Measurement Clusterization}
Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization_illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are identical. Following
\cite{Ucinski05}, we refer to this intriguing phenomenon as
\emph{measurement clusterization}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

@@

Clusterization should not be confused with replication. Replication
requires that the experimentalist executes multiple trials under
circumstances that are \emph{nominally identical} \cite[Section
  1.2.4]{morris2011}. Replication is commonly viewed as a beneficial
and even necessary aspect of optimal experimental design
\cite{fisher1949design, morris2011, schafer2001replication}.
For example, \cite{fisher1949design}, suggested repeating his famous
milk and tea experiment in order "to be able to demonstrate the
predominance of correct classifications in spite of occasional
errors". Unfortunately, in the experiments we consider, replication is
impossible. For example, in the MRI problem, we cannot generate an
individual nominally identical to the one we wish to scan.

Similarly to a design implementing replication, a clustered design
reduces the signal-to-noise ratio of the repeated measurements
\cite{telford2007brief}. The difference is that a clustered design
takes repeated measurements at the expense of other quantities not
measured at all. For example, consider an experiment measuring the
effect of rainfall on grass growth \cite{fay2000rainfall}. The
experiment involved four rainfall manipulation "treatments"
(i.e.~simulating different timing and quantity of rainfall), each
replicated three times over different plots of land. Indeed, it seems
reasonable for researchers to replicate the phenomenon they are trying
to study. A clustered design in such an experiment would imply the
researchers should take a repeated measurement \emph{on the same
plot}, at the expense of measuring grass growth in other plots!

To conclude our short discussion on replication vs.~clusterization:
these are fundamentally different concepts and even though replication
is quite intuitive, it is inapplicable to the inverse problems we
consider in this manuscript.

Researchers of inverse problems widely agree that measurement
clusterization is undesirable \cite{fedorov1996, nyberg2012,
  fedorov1997, Ucinski05, neitzel2019sparse}, prompting the
exploration of various remedies to address this issue. One approach
involves merging close measurements \cite{fedorov1997}; however, this
strategy merely overlooks the phenomenon of measurement
clusterization. An alternative solution lies in
\emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{nyberg2012}.

@@

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
We give two complementing answers to this question. First, from a
  theoretical perspective, we show that clusterization mainly depends
  on how quickly the eigenvalues of the prior covariance in
  observation space decay.
See Section \ref{section:vanishing} --- particularly, Theorem
  \ref{thm:char} and the discussion following it. Furthermore, in
  Section \ref{subsec:lemma_sims} we show results of numerical
  experiments, where simulations of our model give rise to D-optimal
  designs that exhibit clusterization with high probability. Thus,
  given the genericity of our model, we expect measurement
  clusterization to be a generic and ubiquitous phenomenon.

@@

  The connection to Carath\'eodory's Theorem also builds on the focus
  of measurement effort to the abovementioned select set of prior
  eigenvectors. This allows us to move from an infinite-dimensional
  setting to finite dimensions, where the conditions of
  Carath\'eodory's Theorem hold. We conclude that a D-optimal design
  arises by weighting a small number of measurements. If the number of
  allowed measurements is larger than the small number dictated by
  Carath\'eodory's Theorem, we have excessive weight on some
  measurements, which can be interpreted as clusterization.

@@

\subsubsection{Implications}\label{subsub:implications}
Our answer to Question \ref{q:generic} implies that encountering
clusterization could be expected in many different problems across
many different scientific fields. Researchers that encounter
clusterization should not be surprised or wary. In Our answer to
Question \ref{q:why}, we explain what our view of the cause of
clusterization is. It appears, the cause is generic: a D-optimal
design reduces uncertainty for a select set of prior covariance
eigenvectors --- those with the most prior uncertainty, i.e.~those the
practitioner cares about the most!

A further implication of the analysis we present is that
clusterization can serve as an evidence to the number of relevant
eigenvectors in the problem. These eigenvectors correspond to the
relevant degrees of freedom in the problem. Thus, clusterization tells
us that we might be able to reduce the complexity of a computational
model, e.g.~by dropping discretization points.

One potential cause of clusterization is our choice of prior. Gaussian
priors, coupled with a Gaussian likelihood and a linear forward
problem give rise to a closed form solution for the posterior via
conjugacy. As we show in Section \ref{subsec:lemma_sims}, such
assumptions generically give rise to clusterization. While an
assumption of a Gaussian likelihood is standard, and rooted in the
central limit theorem, an assumption of Gaussian prior is merely a
matter of convenience. Therefore, we advise any practitioner who
encounters clusterization to replace their Gaussian priors with
non-Gaussian priors instead \cite{hosseini2017, hosseini2019}.

Another potential cause of clusterization is linearization. While in a
real applications the model is not necessarily linear, some authors
consider a linearized version of their model when seeking optimal
designs \cite{fedorov1996, neitzel2019sparse}. Our analysis then shows
that the linearity of the forward problem is also an important
ingredient in giving rise to clustered designs. Therefore, our advice
to practitioners is to avoid linearization, and find D-optimal designs
via other methods, e.g.~the sampling method of \cite{ryan2003}.

Given the above discussion, it is our belief that the methods
suggested by other authors to avoid clusterization are merely
overlooking the problem. We do not view clustered designs as
inherently bad on their own. Rather, we suggest that if a clustered
design arises, a practitioner should revisit their modeling choices
--- specifically, their choice of prior and model linearization. If
the practitioner is confident in their analysis, then clustered
designs should be avoided only to the extent necessitated by the
physical measuring apparatus.

Clusterization in D-optimal designs raises the question of whether
D-optimal designs should be pursued at all. While other authors have
established convergence results for decaying measurement error
\cite{knapik2011}, space-filling measurements \cite{teckentrup2020}
and randomly sampled designs \cite{nickl2023}, their results do not
hold for D-optimal designs. Luckily, we can expect convergence from
D-optimal designs (including clustered designs) in the following
sense: the posterior uncertainty ellipsoid will contract to zero along
each one of its eigenvectors. See Section \ref{subsub:convergence} for
a precise statement and a proof based on Theorem \ref{thm:char} ---
the main theorem of this manuscript.

\subsubsection{Other Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the space of \emph{observations}
This is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. Specifically, in its
current form, our model does not allow point evaluations. Thus, while
our model is generic enough to be analytically tractable, one may
argue that our model is too far removed from any real application. To
these claims I would answer that scientists have a long history of
studying models that are bare-bones simplifications of real systems,
e.g.~the Ising model \cite{cipra1987}, the Lorenz system \cite{brin},
the Lotka-Volterra equations \cite{logan2006}, the Carnot engine
\cite{kardar2007} and many others.

 \section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over Hilbert
spaces. Since we are ultimately interested in inferring a function
over some domain, we will keep in mind that these Hilbert spaces
should really be thought of as function spaces, utilizing a setup
similar to \cite{knapik2011}. A deeper treatment of the foundations
for inverse problems over function spaces can be found in
\cite{Stuart10}.

\subsection{Forward Problems}\label{subsec:abstract_OED}
In this section we give definitions and notations for forward
problems, which are an essential part of the inverse problems we
discuss later. Consider a "parameter space" \(\hilp\) and an
"observation space" \(\hilo\) --- both separable Hilbert spaces (the
subscripts p and o are for "parameter" and "observation",
respectively). The parameter space includes the quantity we seek to
infer; in the inverse problem of the heat equation, the parameter
space $\hilp$ is where the initial condition lives. The observation
space $\hilo$, on the other hand, is the space from which we take
measurements; in the example of the heat equation, $u(\cdot, T) \in
\hilo$.

@@

Proposition \ref{prop:bigger_better} implies that it is a good idea to
bound the norm of measurements. If, for example, we can take
measurements in $\textup{span}\{\meas\}$ for some $\meas \neq 0$, then
the D-optimality criterion is unbounded, so a D-optimal design does
not exist. In contrast, in any real-life problem where sensors are
concerned, the norm of measurements recorded by sensors is always
one\footnote{Again, our analysis does not directly apply to point
evaluations. We just utilize point evaluations for motivation. We can
approximate point evaluations by e.g.~elements in $\hilo^*$ as long as
$\hilo$ is a function space, e.g.~$L^2(\Omega)$. In this case, for a
fixed apporximation of $\delta$ the norm of the corresponding
functional is a constant $\neq 1$.}:
